---
title: "Ziqi Zhao CSCI E-63C Week 13 Problem Set"
output:
  html_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
ptStart <- proc.time()
library(neuralnet)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(grid)
plot.nn <-
function (x, rep = NULL, x.entry = NULL, x.out = NULL, radius = 0.15, 
    arrow.length = 0.2, intercept = TRUE, intercept.factor = 0.4, 
    information = TRUE, information.pos = 0.1, col.entry.synapse = "black", 
    col.entry = "black", col.hidden = "black", col.hidden.synapse = "black", 
    col.out = "black", col.out.synapse = "black", col.intercept = "blue", 
    fontsize = 12, dimension = 6, show.weights = TRUE, file = NULL, 
    ...) 
{
    net <- x
    if (is.null(net$weights)) 
        stop("weights were not calculated")
    if (!is.null(file) && !is.character(file)) 
        stop("'file' must be a string")
    if (is.null(rep)) {
        for (i in 1:length(net$weights)) {
            if (!is.null(file)) 
                file.rep <- paste(file, ".", i, sep = "")
            else file.rep <- NULL
            #dev.new()
            plot.nn(net, rep = i, x.entry, x.out, radius, arrow.length, 
                intercept, intercept.factor, information, information.pos, 
                col.entry.synapse, col.entry, col.hidden, col.hidden.synapse, 
                col.out, col.out.synapse, col.intercept, fontsize, 
                dimension, show.weights, file.rep, ...)
        }
    }
    else {
        if (is.character(file) && file.exists(file)) 
            stop(sprintf("%s already exists", sQuote(file)))
        result.matrix <- t(net$result.matrix)
        if (rep == "best") 
            rep <- as.integer(which.min(result.matrix[, "error"]))
        if (rep > length(net$weights)) 
            stop("'rep' does not exist")
        weights <- net$weights[[rep]]
        if (is.null(x.entry)) 
            x.entry <- 0.5 - (arrow.length/2) * length(weights)
        if (is.null(x.out)) 
            x.out <- 0.5 + (arrow.length/2) * length(weights)
        width <- max(x.out - x.entry + 0.2, 0.8) * 8
        radius <- radius/dimension
        entry.label <- net$model.list$variables
        out.label <- net$model.list$response
        neuron.count <- array(0, length(weights) + 1)
        neuron.count[1] <- nrow(weights[[1]]) - 1
        neuron.count[2] <- ncol(weights[[1]])
        x.position <- array(0, length(weights) + 1)
        x.position[1] <- x.entry
        x.position[length(weights) + 1] <- x.out
        if (length(weights) > 1) 
            for (i in 2:length(weights)) {
                neuron.count[i + 1] <- ncol(weights[[i]])
                x.position[i] <- x.entry + (i - 1) * (x.out - 
                  x.entry)/length(weights)
            }
        y.step <- 1/(neuron.count + 1)
        y.position <- array(0, length(weights) + 1)
        y.intercept <- 1 - 2 * radius
        information.pos <- min(min(y.step) - 0.1, 0.2)
        if (length(entry.label) != neuron.count[1]) {
            if (length(entry.label) < neuron.count[1]) {
                tmp <- NULL
                for (i in 1:(neuron.count[1] - length(entry.label))) {
                  tmp <- c(tmp, "no name")
                }
                entry.label <- c(entry.label, tmp)
            }
        }
        if (length(out.label) != neuron.count[length(neuron.count)]) {
            if (length(out.label) < neuron.count[length(neuron.count)]) {
                tmp <- NULL
                for (i in 1:(neuron.count[length(neuron.count)] - 
                  length(out.label))) {
                  tmp <- c(tmp, "no name")
                }
                out.label <- c(out.label, tmp)
            }
        }
        grid.newpage()
        pushViewport(viewport(name = "plot.area", width = unit(dimension, 
            "inches"), height = unit(dimension, "inches")))
        for (k in 1:length(weights)) {
            for (i in 1:neuron.count[k]) {
                y.position[k] <- y.position[k] + y.step[k]
                y.tmp <- 0
                for (j in 1:neuron.count[k + 1]) {
                  y.tmp <- y.tmp + y.step[k + 1]
                  result <- calculate.delta(c(x.position[k], 
                    x.position[k + 1]), c(y.position[k], y.tmp), 
                    radius)
                  x <- c(x.position[k], x.position[k + 1] - result[1])
                  y <- c(y.position[k], y.tmp + result[2])
                  grid.lines(x = x, y = y, arrow = arrow(length = unit(0.15, 
                    "cm"), type = "closed"), gp = gpar(fill = col.hidden.synapse, 
                    col = col.hidden.synapse, ...))
                  if (show.weights) 
                    draw.text(label = weights[[k]][neuron.count[k] - 
                      i + 2, neuron.count[k + 1] - j + 1], x = c(x.position[k], 
                      x.position[k + 1]), y = c(y.position[k], 
                      y.tmp), xy.null = 1.25 * result, color = col.hidden.synapse, 
                      fontsize = fontsize - 2, ...)
                }
                if (k == 1) {
                  grid.lines(x = c((x.position[1] - arrow.length), 
                    x.position[1] - radius), y = y.position[k], 
                    arrow = arrow(length = unit(0.15, "cm"), 
                      type = "closed"), gp = gpar(fill = col.entry.synapse, 
                      col = col.entry.synapse, ...))
                  draw.text(label = entry.label[(neuron.count[1] + 
                    1) - i], x = c((x.position - arrow.length), 
                    x.position[1] - radius), y = c(y.position[k], 
                    y.position[k]), xy.null = c(0, 0), color = col.entry.synapse, 
                    fontsize = fontsize, ...)
                  grid.circle(x = x.position[k], y = y.position[k], 
                    r = radius, gp = gpar(fill = "white", col = col.entry, 
                      ...))
                }
                else {
                  grid.circle(x = x.position[k], y = y.position[k], 
                    r = radius, gp = gpar(fill = "white", col = col.hidden, 
                      ...))
                }
            }
        }
        out <- length(neuron.count)
        for (i in 1:neuron.count[out]) {
            y.position[out] <- y.position[out] + y.step[out]
            grid.lines(x = c(x.position[out] + radius, x.position[out] + 
                arrow.length), y = y.position[out], arrow = arrow(length = unit(0.15, 
                "cm"), type = "closed"), gp = gpar(fill = col.out.synapse, 
                col = col.out.synapse, ...))
            draw.text(label = out.label[(neuron.count[out] + 
                1) - i], x = c((x.position[out] + radius), x.position[out] + 
                arrow.length), y = c(y.position[out], y.position[out]), 
                xy.null = c(0, 0), color = col.out.synapse, fontsize = fontsize, 
                ...)
            grid.circle(x = x.position[out], y = y.position[out], 
                r = radius, gp = gpar(fill = "white", col = col.out, 
                  ...))
        }
        if (intercept) {
            for (k in 1:length(weights)) {
                y.tmp <- 0
                x.intercept <- (x.position[k + 1] - x.position[k]) * 
                  intercept.factor + x.position[k]
                for (i in 1:neuron.count[k + 1]) {
                  y.tmp <- y.tmp + y.step[k + 1]
                  result <- calculate.delta(c(x.intercept, x.position[k + 
                    1]), c(y.intercept, y.tmp), radius)
                  x <- c(x.intercept, x.position[k + 1] - result[1])
                  y <- c(y.intercept, y.tmp + result[2])
                  grid.lines(x = x, y = y, arrow = arrow(length = unit(0.15, 
                    "cm"), type = "closed"), gp = gpar(fill = col.intercept, 
                    col = col.intercept, ...))
                  xy.null <- cbind(x.position[k + 1] - x.intercept - 
                    2 * result[1], -(y.tmp - y.intercept + 2 * 
                    result[2]))
                  if (show.weights) 
                    draw.text(label = weights[[k]][1, neuron.count[k + 
                      1] - i + 1], x = c(x.intercept, x.position[k + 
                      1]), y = c(y.intercept, y.tmp), xy.null = xy.null, 
                      color = col.intercept, alignment = c("right", 
                        "bottom"), fontsize = fontsize - 2, ...)
                }
                grid.circle(x = x.intercept, y = y.intercept, 
                  r = radius, gp = gpar(fill = "white", col = col.intercept, 
                    ...))
                grid.text(1, x = x.intercept, y = y.intercept, 
                  gp = gpar(col = col.intercept, ...))
            }
        }
        if (information) 
            grid.text(paste("Error: ", round(result.matrix[rep, 
                "error"], 6), "   Steps: ", result.matrix[rep, 
                "steps"], sep = ""), x = 0.5, y = information.pos, 
                just = "bottom", gp = gpar(fontsize = fontsize + 
                  2, ...))
        popViewport()
        if (!is.null(file)) {
            weight.plot <- recordPlot()
            save(weight.plot, file = file)
        }
    }
}
calculate.delta <-
function (x, y, r) 
{
    delta.x <- x[2] - x[1]
    delta.y <- y[2] - y[1]
    x.null <- r/sqrt(delta.x^2 + delta.y^2) * delta.x
    if (y[1] < y[2]) 
        y.null <- -sqrt(r^2 - x.null^2)
    else if (y[1] > y[2]) 
        y.null <- sqrt(r^2 - x.null^2)
    else y.null <- 0
    c(x.null, y.null)
}
draw.text <-
function (label, x, y, xy.null = c(0, 0), color, alignment = c("left", 
    "bottom"), ...) 
{
    x.label <- x[1] + xy.null[1]
    y.label <- y[1] - xy.null[2]
    x.delta <- x[2] - x[1]
    y.delta <- y[2] - y[1]
    angle = atan(y.delta/x.delta) * (180/pi)
    if (angle < 0) 
        angle <- angle + 0
    else if (angle > 0) 
        angle <- angle - 0
    if (is.numeric(label)) 
        label <- round(label, 5)
    pushViewport(viewport(x = x.label, y = y.label, width = 0, 
        height = , angle = angle, name = "vp1", just = alignment))
    grid.text(label, x = 0, y = unit(0.75, "mm"), just = alignment, 
        gp = gpar(col = color, ...))
    popViewport()
}
```

# Preface

The goal of this problem set is to develop some intuition about the impact of the number of nodes in the hidden layer of the neural network.  We will use few simulated examples to have clear understanding of the structure of the data we are modeling and will assess how performance of the neural network model is impacted by the structure in the data and the setup of the network.

First of all, to compensate for lack of coverage on this topic in ISLR, let's go over a couple of simple examples.  We start with simulating a simple two class dataset in 2D predictor space with an outcome representative of an interaction between attributes.  (Please notice that for the problems you will be working on this week you will be asked below to simulate a dataset using a different model.)

```{r, fig.height=7,fig.width=7}
# fix seed so that narrative always matches the plots:
set.seed(1234567890)
nObs <- 1000
ctrPos <- 2
xyTmp <- matrix(rnorm(4*nObs),ncol=2)
xyCtrsTmp <- matrix(sample(c(-1,1)*ctrPos,nObs*4,replace=TRUE),ncol=2)
xyTmp <- xyTmp + xyCtrsTmp
gTmp <- paste0("class",(1+sign(apply(xyCtrsTmp,1,prod)))/2)
plot(xyTmp,col=as.numeric(factor(gTmp)),pch=as.numeric(factor(gTmp)),xlab="X1",ylab="X2")
abline(h=0)
abline(v=0)
```

Symbol color and shape indicate class.  Typical problem that will present a problem for any approach estimating a single linear decision boundary.  We used similar simulated data for the random forest (week 10) problem set.

## One hidden node

We can fit simple neural network (using all default values in the call to `neuralnet` -- notice that both covariates and outcome have to be numeric as opposed to factor) and plot its layout (allowing for its output to be included in Rmarkdown generated report actually seems to be quite painful - one has to overwrite original implementation of `plot.nn` with the one that doesn't call `dev.new()` that is included in this Rmarkdown file with `echo=FALSE` -- to do the same you have to include that block into your Rmarkdown file also):

```{r,fig.height=7,fig.width=7}
### Doesn't run: "requires numeric/complex ... arguments"
### nnRes <- neuralnet(g~X1+X2,data.frame(g=gTmp,xyTmp))
nnRes <- neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp)),xyTmp))
plot(nnRes)
```

That shows us a model with one node in a single hidden layer (default parameters).

We can lookup actual model predictions and recalculate them from input variables (in the field `covariate`) and model weight and activation function (fields `weights` and `act.fct` respectively):

```{r}
head(nnRes$net.result[[1]])
cbind(rep(1,6),nnRes$act.fct(cbind(rep(1,6),nnRes$covariate[1:6,])%*%nnRes$weights[[1]][[1]]))%*%nnRes$weights[[1]][[2]]
```

Notice that input parameter `linear.output` governs whether activation function is called on the value of the output node or not:

```{r}
nnResNLO <- neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp)),xyTmp),linear.output=FALSE)
head(nnResNLO$net.result[[1]])
nnResNLO$act.fct(cbind(rep(1,6),nnResNLO$act.fct(cbind(rep(1,6),nnResNLO$covariate[1:6,])%*%nnResNLO$weights[[1]][[1]]))%*%nnResNLO$weights[[1]][[2]])
quantile(nnResNLO$net.result[[1]])
```

As the last statement (the quantiles of the predicted out) above shows, the use of activation function limiting predicted values to $[0;1]$ range when modeling outcome taking values outside of $[0;1]$ interval does not result in a very useful model. In this case with true outcome values constrained to $\{1,2\}$ so that the error is minimized by predicting every outcome to be as close to $1$ as possible.

Using binary -- 0 or 1 -- outcome produces more useful model when activation function is applied to the output node (`linear.output=FALSE`) and allows use of cross-entropy error function (often used in classification setting in combination with the activation function applied to the output layer):

```{r}
nnResNLO01 <- neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp))-1,xyTmp),linear.output=FALSE)
quantile(nnResNLO01$net.result[[1]],c(0,0.1,0.25,0.5,1))
if ( FALSE ) {
  # from v.3.5.x(?) this started to throw an error instead of silent switch to "sse":
  nnResNLO12CE <- neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp)),xyTmp),linear.output=FALSE,err.fct="ce")
}
nnResNLO01CE <- neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp))-1,xyTmp),linear.output=FALSE,err.fct="ce")
head(nnResNLO01CE$net.result[[1]])
nnResNLO01CE$act.fct(cbind(rep(1,6),nnResNLO01CE$act.fct(cbind(rep(1,6),nnResNLO01CE$covariate[1:6,])%*%nnResNLO01CE$weights[[1]][[1]]))%*%nnResNLO01CE$weights[[1]][[2]])
quantile(nnResNLO01CE$net.result[[1]])
```

We can plot model output indicating class identity (left panel below) that tells us that when true outcome values are constrained to 1 or 2, sum of squared errors is used as error function and output node values are used as-is (not transformed by activation function -- `linear.output=TRUE` by default), the majority of the points were estimated to be close to 1 or 1.6 and that majority of those estimated to be close to 1 correspond to the about half of the observations at the first level of the class factor (i.e. numerical value of 1).  It is also easy to see that those with predicted value of 1.6 represent roughly 1:2 mix of observations from the first and second levels of the outcome respectively, so that $1.6 \approx (1+2*2)/3 = 5/3$ approximately equals average of their numerical values corresponding to the levels of the factor representing them. 

The nature of the model estimated by `neuralnet` in this (very simple!) case becomes even more intuitive if we render all points in the area encompassing our training set with model predictions and overlay training dataset on top of that (right panel below).  It is immediately apparent that this model identified a line in this 2D space separating one cloud of points belonging mostly to one class from all others so that predicted values are approximately equal to the average outcome on each side of this decision boundary.

```{r,fig.height=6,fig.width=12}
plotNNpreds2D2class <- function(inpNN,inpClassThresh,inpGrid=(-60:60)/10) {
  tmpClrPch <- as.numeric(factor(inpNN$response))
  plot(inpNN$net.result[[1]],col=tmpClrPch,pch=tmpClrPch)
  table(inpNN$net.result[[1]][,1]>inpClassThresh,inpNN$response)
  xyGridTmp <- cbind(X1=rep(inpGrid,length(inpGrid)),X2=sort(rep(inpGrid,length(inpGrid))))
  # before predict.nn existed:
  #gridValsTmp <- compute(inpNN,xyGridTmp)$net.result
  gridValsTmp <- predict(inpNN,xyGridTmp)
  errTmp <- sum(inpNN$err.fct(inpNN$net.result[[1]][,1],inpNN$response))
  plot(xyGridTmp,col=as.numeric(gridValsTmp>inpClassThresh)+1,pch=20,cex=0.3,main=paste("Error:",round(errTmp,6)))
  points(inpNN$covariate,col=tmpClrPch,pch=tmpClrPch)
  ## Equations defining decision boundary:
  ## 1*w0 + X1*w1 + X2*w2 = 0, i.e.:
  ## 0 = inpNN$weights[[1]][1]+inpNN$weights[[1]][2]*X1+inpNN$weights[[1]][3]*X2, i.e:
  ## X2 = (-inpNN$weights[[1]][1] - inpNN$weights[[1]][2]*X1) / inpNN$weights[[1]][3]
  for ( iTmp in 1:ncol(inpNN$weights[[1]][[1]]) ) {
    abline(-inpNN$weights[[1]][[1]][1,iTmp] / inpNN$weights[[1]][[1]][3,iTmp], -inpNN$weights[[1]][[1]][2,iTmp] /inpNN$weights[[1]][[1]][3,iTmp],lty=2,lwd=2)
  }
}
old.par <- par(mfrow=c(1,2),ps=16)
plotNNpreds2D2class(nnRes,1.3)
par(old.par)
```

Similar, aside from a different position of decision boundary, results can be obtained when using binary representation of the outcome with cross-entropy as error function together with applying activation function to the output layer:

```{r,fig.width=8,fig.height=8}
plot(nnResNLO01CE)
```

Once activation function is applied to the output node, the output values are bound to the $[0,1]$ interval.  The predicted values that are far enough from the decision boundary are also approximately set to the average of the outcome in that subspace:

```{r,fig.height=6,fig.width=12}
old.par <- par(mfrow=c(1,2),ps=16)
plotNNpreds2D2class(nnResNLO01CE,0.5)
par(old.par)
```

The important points resulting from the results shown in the figures above are the following:

* as simple of a model as the one that was employed here (with one node in a single hidden layer along with all other default parameters) cannot do much better than what we observed here
* because calling (default - logistic) activation function on a given linear combination of the input variables more or less amounts to assigning almost all points on one side of hyperplane (line in 2D, plane in 3D, etc.) to zero and on the other side -- to unity
* weights involved in transforming outcome of the hidden layer into model predictions will change those zeroes and ones to values closer to the desired outcome values, but still, use of such a simple model (with a single hidden node) employed here to prime our intuition more or less amounts to splitting covariate space into two half spaces by a hyperplane and assigning almost constant outcomes to the vast majority of the points on either side of it
* the weights for the inputs into the single hidden node shown in the network layout plot above and stored in `weights` field of the result returned by `neuralnet` define this hyperplane (line in 2D, etc.) shown as dashes in the panels on the right above
* this hyperplane is where sum of weighted input variables and an intercept is identical zero (and thus the result of logistic activation function is 0.5 rapidly becoming zero or one for points further away from this boundary)

## Two hidden nodes, single hidden layer

Now, let's add another node to the hidden layer of this network.  From the above, we know what to expect as a result of that -- another hyperplane (line in 2D) will be added to the space of covariates now dividing it into (depending on whether those hyperplans are almost parallel or not) three or four subspaces, consequently, assigning most of the points to three or four potentially different constants.  Clearly, this level of granularity could suffice for developing a model that would do quite well in our toy example.

To have more than one node in the hidden layer we set `hidden` parameter to the number of nodes in it (length of vector provided as `hidden` parameter governs the number of the hidden *layers* in the network -- we still use one layer here):

```{r,fig.height=7,fig.width=7}
set.seed(1234567)
nnRes2 <- neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp)),xyTmp),hidden=2)
plot(nnRes2)
```

We can see that now resulting network has two nodes in a single hidden layer, which two covariates enter with weights that are approximately comparable in magnitude and opposite in sign.  Their comparably weighted sum added to a constant close to one gives the outcome value of this model. The effect of those weights in defining decision boundaries in the space of predictors is best seen from the figure below:

```{r,fig.height=6,fig.width=12}
old.par <- par(mfrow=c(1,2),ps=16)
plotNNpreds2D2class(nnRes2,1.5)
par(old.par)
```

This model sets up two almost parallel lines that encompass most of the observations from the first class, leaving most of the observations from the second class outside of the resulting slab.  Now let's repeat fitting neural network three times (each time starting with random choice of starting weights in the model) and compare stability of the resulting models:

```{r,fig.height=6,fig.width=9}
old.par <- par(mfcol=c(2,3),ps=16)
for ( iTry in 1:3 ) {
  nnRes2 <- neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp)),xyTmp),hidden=2)
  plotNNpreds2D2class(nnRes2,1.5)
}
par(old.par)
```

We can see that quite frequently given the parameters used the process converges to a suboptimal solution with about half of the observations remaining in "gray" zone where their assignment to either of the classes is not immediately apparent.

Aside from the multitude of local minima for neural network fitting procedure that could prevent it from finding better solutions, the main point to take from this exercise is that adding more nodes to the hidden layer (with all other default choices employed here) amounts to adding more hyperplanes bisecting the space of predictors, creating more and more subspaces where the outcome can take different values (often close to a constant in each subspace).  Obviously, the geometry of the resulting decision surfaces can become quite complicated even with modest number of nodes in the hidden layer. Lastly, these considerations provide some intuition for considering what could be a useful number of hidden nodes in the model. In thinking about that it might be useful to consider how many such hyperplanes could be sufficient to effectively separate observations belonging to different outcome categories.  Not that we necessarily would have such knowledge ahead of time, but this might prove to be a complementary way to think about the problem in addition to the often sited empirical guidelines that are based on the number of predictor variables, etc.

The point of this problem set is to assess how these aspects of neural network fitting play out in another simulated dataset.

Lastly, we also would like you to consider what combination of error function and output node transformation you would like to use for this week problem set.  Below are three calls of `neuralnet` timed using different combinations of `err.fct` and `linear.output`:

```{r}
system.time(invisible(neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp))-1,xyTmp),hidden=2,linear.output = TRUE,err.fct="sse")))
system.time(invisible(neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp))-1,xyTmp),hidden=2,linear.output = FALSE,err.fct="sse")))
system.time(invisible(neuralnet(g~X1+X2,data.frame(g=as.numeric(factor(gTmp))-1,xyTmp),hidden=2,linear.output = FALSE,err.fct="ce")))
```

Obviously, for our toy example use of untransformed outcome from the output node and sum of squares as error function results in the fastest convergence.  But then the decision as to how to translate its predictions to class categories is yours.  On the other hand, one might argue that more principled approach since we are dealing with classification problem would be to use logistic transform of the outcome to contain it within $[0;1]$ interval and cross-entropy as an error function.  Except that it seems to run nearly two orders of magnitude slower in this case and does not necessarily converge with default values for the rest of the arguments.  Please feel free to experiment with how they will perform for the problems you are presented with below and decide for yourself which combination of these parameters is more suitable for the task at hand.

# Problems

## Problem 1: generate 3D data with cube as a decision surface (points: 10/20 – graduate/undergraduate)

Simulate data with n=1000 observations and p=3 covariates -- all random values from standard normal distribution ($\mathcal{N}\left(0,1\right), \mu=0,\sigma=1$).  Create two category class variable assigning all observations within a cube centered at $(0,0,0)$ with the edge length of $2.5$ (i.e. with vertices at $(1.25,1.25,1.25)$, $(1.25,1.25,-1.25)$, ..., $(-1.25,-1.25,-1.25)$) -- to one class category and observations outside of this cube -- to the second class category. Confirm that the simulated observations are nearly evenly split between the two classes.

Please note that this dataset is entirely different from the one used in the preface -- you will need to write code simulating it on your own.  Somewhat related 2D dataset was used as a motivational example at week 12 (SVM) lecture before introducing kernels and SVMs. However, the example in the lecture was in 2D (three-dimensional problem here) and had a circular (in 3D -- spherical) boundary (here we work with a cube as a decision surface).  Since you will be reusing this code in the following two problems it is probably best to turn this procedure into a function with appropriate parameters.

Ten points available for this problem are composed of accomplishing the following tasks:

1. Correct implementation of the function generating data as prescribed above (3 points)
2. Check and demonstrate numerically that the resulting class assignment splits these observations, subject to sampling variability, evenly between these two groups (3 points)
3. Plot values of the resulting covariates projected at each pair of the axes indicating classes to which observations belong with symbol color and/or shape (you can use function `pairs`, for example) (2 points)
4. Reflect on the geometry of this problem by answering the following question: what is the smallest number of planes in 3D space that would completely enclose points from the "inner" class?  Is this number equal to the number of cube faces or is it something smaller? Larger?  Please note that "enclose" above does *not* mean "perfectly discriminate between the points assigned to two classes" (2 points) 

```{R P1}
# Function to generate 3D data with cube as decision surface
generate_3D_data <- function(n = 1000, edge_length = 2.5) {
  # Generate random data from standard normal distribution
  data <- data.frame(X1 = rnorm(n), X2 = rnorm(n), X3 = rnorm(n))
  
  # Define the class variable, 1 for outer, 0 for inner
  half_edge <- edge_length / 2
  data$class <- ifelse(abs(data$X1) <= half_edge & abs(data$X2) <= half_edge & abs(data$X3) <= half_edge, 0, 1)
  
  return(data)
}

# Generate data
cube_data <- generate_3D_data()

# Check class distribution
class_distribution <- table(cube_data$class)
print(class_distribution)

# Plot projections of the covariates
pairs(cube_data[, 1:3], col = ifelse(cube_data$class == 0, "red", "blue"), pch = 19)

```

**Discussions:**

Reflecting on geometry, the smallest number of planes to enclose points from the inner class is 4 faces in a tetrahedron shape, as a large enough tetrahedron can enclose any number of points less than infinity. However, a tetrahedron will likely neither do this efficiently with space, nor discriminate between the points of the two classes, 6 faces would do much betterr as the cube shape has 6 faces.


### Extra points problem: boundary for equal split (points: 5/5 – graduate/undergraduate)

The boundary for assigning observations to the inner class used above -- $\max_i |X_i| \leq 1.25,i=\{1,2,3\}$ -- splits observations from standard normal in 3D nearly evenly between the two classes, but not quite exactly 50/50. Please derive mathematical expression for the value to be used instead of 1.25 in the expression above to contain *exactly* half of probability density within the inner cube, present its numerical value to the 9-th decimal place and demonstrate numerically that it results in closer to equal split of the simulated observation between the two classes than $1.25$.

Please note that this is *not* a problem of devising an *algorithm* equally splitting a given dataset in two, but purely probabilistic/mathematical question -- what threshold to use for a cube size above that will split such dataset *exactly* evenly on average / in the limit of infinitely large sample size (i.e. that contains precisely half of the probability density of standard normal distribution in 3D).


```{R P1.2}

# Function to find equal split threshold
find_equal_split_threshold <- function() {
  # Define function to calculate volume of the cube that falls within the threshold
  volume_within_cube <- function(threshold) {
    # Volume = P(-threshold <= X1 <= threshold) * P(-threshold <= X2 <= threshold) * P(-threshold <= X3 <= threshold)
    p <- pnorm(threshold) - pnorm(-threshold)
    return(p^3)
  }
  
  # Binary search to find the threshold that results in 50% probability
  target <- 0.5
  low <- 0
  high <- 3
  tol <- 1e-9
  
  while ((high - low) > tol) {
    mid <- (low + high) / 2
    if (volume_within_cube(mid) > target) {
      high <- mid
    } else {
      low <- mid
    }
  }
  
  return((low + high) / 2)
}

# Find the threshold for equal split
equal_split_threshold <- find_equal_split_threshold()
print(paste("Equal split threshold:", equal_split_threshold))

# Validate the threshold numerically
cube_data_equal_split <- generate_3D_data(edge_length = 2 * equal_split_threshold)
equal_split_distribution <- table(cube_data_equal_split$class)
print(equal_split_distribution)
```
**Discussions:**

Mathematically, the probability content of the cube can be expressed as a triple integral (A little hard to show in R markdown). Then, the integral in one dimension is the cumulative distribution function (CDF) of the standard normal distribution and we can take it cubed, thus:

1. P(inside cube)=[2Φ(c)−1]^3
2. To split the observations evenly, we would need P(inside cube)=0.5

Let 1 equal to 2 here, we get:

[2Φ(c)−1]^3 = 0.5

Then solving:

Φ(c)= (1+0.5^(1/3))/2

Finally we have to use inverse CDF of standard normal distribution and solve to obtain c:
 
c=1.263806538

The above code also confirms this number.
 

## Problem 2: neural network classifier (points: 20/40 – graduate/undergraduate)

For the dataset simulated above fit neural networks with 1 through 6 nodes in a single hidden layer (use `neuralnet` implementation).  For each of them calculate training error (see an example in Preface where it was calculated using `err.fct` field in the result returned by `neuralnet`).  Simulate another independent dataset (with n=10,000 observations to make resulting test error estimates less variable) using the same procedure as above (3D standard normal, two classes, cube as a decision surface) and use it to calculate the test error at each number of hidden nodes.  Plot training and test errors as function of the number of nodes in the hidden layer.  What does resulting plot tells you about the interplay between model error, model complexity and problem geometry?  What is the geometrical interpretation of this error behavior?


```{R P2 using ce}

# Generate datasets
train_data <- generate_3D_data(n = 200)
test_data <- generate_3D_data(n = 10000)

# Initialize vectors for errors
train_errors <- numeric(6)
test_errors <- numeric(6)

# Loop through hidden nodes
for (hidden_nodes in 1:6) {
  cat("\nTraining model with", hidden_nodes, "hidden nodes...\n")
  
  nn_model <- neuralnet(
    class ~ X1 + X2 + X3,
    data = train_data,
    hidden = hidden_nodes,
    linear.output = FALSE,
    err.fct = "ce",
    stepmax = 1e+07  # Increased stepmax
  )
  
  # Compute predictions for training data
  train_pred <- compute(nn_model, train_data[, c("X1", "X2", "X3")])$net.result
  train_pred <- ifelse(train_pred > 0.5, 1, 0)  # Convert probabilities to binary class labels

  # Calculate train classification error (proportion of misclassified examples)
  train_error <- mean(train_pred != train_data$class)
  train_errors[hidden_nodes] <- train_error
  
  # Compute test predictions
  test_pred <- compute(nn_model, test_data[, c("X1", "X2", "X3")])$net.result
  test_pred <- ifelse(test_pred > 0.5, 1, 0)
  
  # Save test error
  test_error <- mean(test_pred != test_data$class)
  test_errors[hidden_nodes] <- test_error
  
  # Print train and test errors for this iteration
  cat("Train error for", hidden_nodes, "hidden nodes:", train_error, "\n")
  cat("Test error for", hidden_nodes, "hidden nodes:", test_error, "\n")

}
 
# Plot training and test errors
plot(1:6, train_errors, type = "b", col = "red", pch = 19, ylim = c(0, max(c(train_errors, test_errors))),
     xlab = "Number of Hidden Nodes", ylab = "Mean Error Rate", main = "Training and Test Errors")
lines(1:6, test_errors, type = "b", col = "blue", pch = 19)
legend("topright", legend = c("Training Error", "Test Error"), col = c("red", "blue"), pch = 19)
```

**Discussions:**

The above uses cross-entropy. It would get stuck running on my local machine (Macbook Pro M3 chip) with n=1000 for the train data, so I had to reduce to n=200.

As the number of nodes in the hidden layer increases, the training error decreases. This is because a more complex model (with more hidden nodes) can better capture the nuances of the training data, reducing underfitting. However，too many nodes likely results in overfitting and the model becomes less robust as we see here, the train vs test error gap increases as nodes increase.

For test error, initially, as the number of nodes increases, the test error decreases due to better learning of the decision boundary. Beyond a certain number of nodes (5 as shown in the plot), the test error starts to increase again as the model overfits the training data and loses generalization. This behavior reflects the bias-variance tradeoff.

The problem geometry is defined by the 3D cube decision surface. However, the cube's sharp edges and flat faces make the decision boundary non-linear and challenging to model. Simpler models may struggle to capture these details, resulting in higher errors.

Geometrically interpreting this:

- With 1 or 2 hidden nodes, the model lacks the capacity to approximate the 3D cube's decision surface effectively. The decision boundary may appear as a simple plane or sphere, failing to capture the cube's sharp edges and corners, leading to higher training and test errors.

- Around 4 to 5 hidden nodes, the model achieves a balance between capacity and generalization. 

- With 5 or 6 hidden nodes, the model becomes overly complex and starts fitting noise in the training data. This leads to sharp fluctuations in the decision boundary, deviating from the true cube shape and increasing test error.


```{R P2 using sse}
# Generate datasets
train_data <- generate_3D_data(n = 1000)
test_data <- generate_3D_data(n = 10000)

# Initialize vectors for errors
train_errors <- numeric(6)
test_errors <- numeric(6)

# Loop through hidden nodes
for (hidden_nodes in 1:6) {
  cat("\nTraining model with", hidden_nodes, "hidden nodes...\n")
  
  nn_model <- neuralnet(
    class ~ X1 + X2 + X3,
    data = train_data,
    hidden = hidden_nodes,
    linear.output = FALSE,
    err.fct = "sse",
    stepmax = 1e+07  # Increased stepmax given error
  )
  
  # Compute predictions for training data
  train_pred <- compute(nn_model, train_data[, c("X1", "X2", "X3")])$net.result
  train_pred <- ifelse(train_pred > 0.5, 1, 0)  # Convert probabilities to binary class labels

  # Calculate train classification error (proportion of misclassified examples)
  train_error <- mean(train_pred != train_data$class)
  train_errors[hidden_nodes] <- train_error
  
  # Compute test predictions
  test_pred <- compute(nn_model, test_data[, c("X1", "X2", "X3")])$net.result
  test_pred <- ifelse(test_pred > 0.5, 1, 0)
  
  # Save test error
  test_error <- mean(test_pred != test_data$class)
  test_errors[hidden_nodes] <- test_error
  
  # Print train and test errors for this iteration
  cat("Train error for", hidden_nodes, "hidden nodes:", train_error, "\n")
  cat("Test error for", hidden_nodes, "hidden nodes:", test_error, "\n")

}
 
# Plot training and test errors
plot(1:6, train_errors, type = "b", col = "red", pch = 19, ylim = c(0, max(c(train_errors, test_errors))),
     xlab = "Number of Hidden Nodes", ylab = "Mean Error Rate", main = "Training and Test Errors")
lines(1:6, test_errors, type = "b", col = "blue", pch = 19)
legend("topright", legend = c("Training Error", "Test Error"), col = c("red", "blue"), pch = 19)

```
The above is the exact same code, but run using sse for comparison, still fairly slow when hidden nodes go above 3 but runnable at n=1000 for the train data.

Comparison:

The above results with sse seems more robust and there is less gap between test and train error. We also don't see the U plot where test error increases again at high enough node count, which happened when using cross entropy. This may be due to the fact that SSE regularizes the model inherently by not emphasizing confidence in predictions as strongly as CE; This can suppress overfitting but may fail to capture subtle variations in the data.

SSE focuses on minimizing squared differences and does not penalize uncertain predictions as strongly as CE. This often results in softer decision boundaries and may lead to better generalization, particularly when the training dataset is small.

## Problem 3: evaluate impacts of sample size and noise (points: 30/30(extra) – graduate/undergraduate)

Setup a simulation repeating procedure described above for n=100, 200 and 500 observations in the *training* set as well as adding none, 1, 2 and 5 null variables to the training and test data (and to the covariates in formula provided to `neuralnet`).  Draw values for null variables from standard normal distribution as well and do not use them in the assignment of the observations to the class category (e.g. `x<-matrix(rnorm(600),ncol=6); cl<-as.numeric(factor(rowSums(abs(x[,1:3])<1.25)==3))` creates dataset with three informative and three null variables). Repeat calculation of training and test errors at least several times for each combination of sample size, number of null variables and size of the hidden layer simulating new training and test dataset every time to assess variability in those estimates.  Present resulting error rates so that the effects of sample size and fraction of null variables can be discerned and discuss their impact of the resulting model fits.  

```{R P3}

# Function to generate datasets with null variables
generate_dataset <- function(n, null_vars) {
  informative_vars <- 3  # Number of informative variables
  total_vars <- informative_vars + null_vars
  x <- matrix(rnorm(n * total_vars), ncol = total_vars)  # Generate data
  colnames(x) <- paste0("X", 1:total_vars)
  
  # Assign class using only informative variables
  class <- as.numeric(rowSums(abs(x[, 1:informative_vars]) < 1.25) == informative_vars)
  
  # Return as a data frame
  data <- as.data.frame(x)
  data$class <- class
  return(data)
}

# Parameters
sample_sizes <- c(100, 200, 500)
null_vars <- c(0, 1, 2, 5)
hidden_nodes <- 1:6
reps <- 5  # Number of repetitions

# Initialize results storage
results <- expand.grid(sample_size = sample_sizes, null_vars = null_vars, hidden_nodes = hidden_nodes, rep = 1:reps)
results$train_error <- NA
results$test_error <- NA

# Simulation
for (i in 1:nrow(results)) {
  cat("\nSimulating for sample size =", results$sample_size[i],
      ", null vars =", results$null_vars[i],
      ", hidden nodes =", results$hidden_nodes[i],
      ", repetition =", results$rep[i], "...\n")
  
  # Generate datasets
  train_data <- generate_dataset(results$sample_size[i], results$null_vars[i])
  test_data <- generate_dataset(10000, results$null_vars[i])  # Large test set
  
  #Using a graphics device that doesn't depend on Quartz (given error)
  options(bitmapType = "cairo")
  
  # Train neural network
  nn_model <- tryCatch(
    neuralnet(
      class ~ .,  # Use all variables in the formula
      data = train_data,
      hidden = results$hidden_nodes[i],
      linear.output = FALSE,
      err.fct = "sse",
      stepmax = 1e+07
    ),
    error = function(e) {
      cat("Error in training model: ", e$message, "\n")
      return(NULL)
    }
  )
  
  if (is.null(nn_model)) next
  
  # Compute training error
  train_pred <- compute(nn_model, train_data[, -ncol(train_data)])$net.result
  train_pred <- ifelse(train_pred > 0.5, 1, 0)
  results$train_error[i] <- mean(train_pred != train_data$class)
  
  # Compute test error
  test_pred <- compute(nn_model, test_data[, -ncol(test_data)])$net.result
  test_pred <- ifelse(test_pred > 0.5, 1, 0)
  results$test_error[i] <- mean(test_pred != test_data$class)
}

options(bitmapType = "quartz")

```

```{r p3.2}
options(bitmapType = "quartz")

# Summarize results
library(dplyr)
summary_results <- results %>%
  group_by(sample_size, null_vars, hidden_nodes) %>%
  summarise(mean_train_error = mean(train_error, na.rm = TRUE),
            sd_train_error = sd(train_error, na.rm = TRUE),
            mean_test_error = mean(test_error, na.rm = TRUE),
            sd_test_error = sd(test_error, na.rm = TRUE)) %>%
  ungroup()

# Visualize results
library(ggplot2)
ggplot(summary_results, aes(x = hidden_nodes, y = mean_test_error, color = as.factor(null_vars))) +
  geom_line() +
  geom_point() +
  facet_wrap(~sample_size, scales = "free_y") +
  labs(
    title = "Impact of Sample Size and Null Variables on Test Error",
    x = "Number of Hidden Nodes",
    y = "Mean Test Error",
    color = "Null Variables"
  ) +
  theme_minimal()
```
**Discussions:**

Impact of sample size: Mean test errors are higher across all configurations of hidden nodes and null variables, seen in n=100 and n=200. We see that the test error decreases with more hidden nodes initially but plateaus at a higher error compared to larger sample sizes. Variability in test error (across null variables) is more pronounced, indicating that small datasets are more susceptible to noise. Larger datasets provide enough information for the model to accurately learn the underlying decision boundary, reducing the impact of noise.

Impact of null variables: Test errors increase as the number of null variables rises. This trend is more pronounced in smaller datasets (n = 100, 200) where the model struggles to differentiate between informative and null features. The null variables act as noise, adding irrelevant dimensions to the input space; The model may allocate weights to null variables, reducing its ability to accurately learn the relationship between informative variables and the target class.

Impact of Hidden Nodes: As shown, for small sample sizes, the optimal number of hidden nodes is typically lower, test errors initially decrease with more hidden nodes but plateau or increase slightly with very high complexity (e.g., 6 hidden nodes). For large sample sizes, the model has sufficient data to avoid overfitting and can benefit from increased complexity. Too little nodes underfit and too many nodes overfit but the threshold increases with a larger sample size.

Finally, note that variability in test error overall increases with smaller datasets and more null variables.


### Extra points problem: model WiFi localization data (points: 10/10 – graduate/undergraduate)

Use `neuralnet` to model the outcome in WiFi localization dataset that we used in previous weeks *using outcome in its original, four-levels format.* Your `neuralnet` models will need to have four output nodes to represent outcome with four levels. Obtain training and test error for this model for several sizes of the hidden layer.  Compare resulting test error in predicting Location 3 to that observed for random forest, SVM and KNN approaches earlier in the course.

I decided not to complete this question this week (Life and Thanksgiving got to me!). 

# Session info  {.unlisted .unnumbered}

For reproducibility purposes it is always a good idea to capture the state of the environment that was used to generate the results:

```{r}
sessionInfo()
```

The time it took to knit this file from beginning to end is about (seconds):

```{r}
proc.time() - ptStart
```
